|

### 2020-07-14 paper.Rmd
Density plot added, but contour lines need to be improved!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(gridExtra)
```

@@

In this work we address the piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05]. Tours show interpolated sequences of low-dimensional projections of the data. When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle. However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points.

Here we suggest to redistribute projected data points such that equal volume in the original high-dimensional space will approximately be mapped onto equal volume in the low-dimensional projection, focussing on equal area in a 2D projection plane. This redistribution is achieved by assuming spherical distributions, and deriving a non-linear transformation of the projected radius. This essentially results in zooming in near the center of the distribution, where the transformation is approximately linear, while points that are further out in the projection get pushed towards the maximum radius.

The paper is structured as follows: the radial transformation is described in Section \ref{sec:method}, the implementation as a new display function for the tourr package is described in Section \ref{sec:implementation}, and we show applications in Section \ref{sec:application}. We conclude in Section \ref{sec:concl}.

XXX see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".

XXX maybe in real applications we need to do both? Use some form of dimension reduction that can screen some of the noise, followed by zooming in on the center?

@@

XXX show density contours, similar to Fig. 3

@@

```{r radii, fig.cap="Relation between $r$ and $r'$ for different values of $p$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
library(tidyverse)
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

@@

```{r circles, fig.cap="Rescaled equidistant concentric circles, for $p=2, 3, 10, 100$. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases."}
library(ggforce)

@@

```{r scaled, eval=F, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$. We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
library(tidyverse)
# define index as function of c
trans_s <- function(s, p){
  p <- s * p
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

@@

XXX this should explain how the display integrates with the tour, in terms of a data pipeline: p-D data -> projection matrix from the tour -> d-D projected data -> radial transformation -> final display
XXX include short tour intro here with relevant references

The sage display has been implemented in R [@rref] in the `display_slice` method in the `tourr` package [@tourr]. This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`. Both these parameters are described in Section \ref{sec:params} above.

@@

Below we show example code demonstrates how the new display redistributes the projected points drawn uniformly inside a 5D hypersphere.
```{r codeExampel, echo=TRUE, eval=FALSE}
library(tourr)
# use geozoo to uniformly generate points inside a 5D sphere
sphere5 <- geozoo::sphere.solid.random(5)$points
colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5") # naming variables
# standard tour animation with default settings
animate_xy(sphere5)
# new sage display to redistribute the points based on projected volume
animate_sage(sphere5)
# we can use R and s to tune the display
# for example use R to zoom in near the center
# we also adjust half_range to fit the display on the plotting canvas
animate_sage(sphere5, R = 0.1, half_range = 0.1)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(gridExtra)
library(ggforce)
library(tourr)
```

@@

XXX FIX - how can I make the contours better? Maybe get radial density and draw as circles?
For illustration we show two-dimensional linear projections of points sampled uniformly within hyperspheres in $p=3, 10, 100$ dimensions in Figure \ref{fig:density}. All distributions are uniform within the maximum radius $R=1$, but the projected points start piling near the center as $p$ increases.

```{r density, fig.cap="Two-dimensional projections of 10k points sampled uniformly within a $p$-dimensional hyperspheres, for $p=3, 10, 100$. The density is shown both through transparancy of the projected points, and as contour lines.", out.width="95%", fig.align = "center"}

lbl <- c("p=3", "p=10", "p=100")
names(lbl) <- c("p3", "p10", "p100")

n <- 10000

# sample points, only keep first two components for 2D projection
p3 <- geozoo::sphere.solid.random(3, n)$points[, c(1,2)]
p10 <- geozoo::sphere.solid.random(10, n)$points[, c(1,2)]
p100 <- geozoo::sphere.solid.random(100, n)$points[, c(1,2)]
colnames(p3) <- c("x", "y")
colnames(p10) <- c("x", "y")
colnames(p100) <- c("x", "y")

proj_points <- as_tibble(rbind(p3, p10, p100)) %>%
  add_column(p = c(rep(3, n), rep(10, n), rep(100, n)))

ggplot(proj_points, aes(x, y)) +
  geom_point(alpha=0.02) + 
  geom_density2d(bins = 7, contour_var = "count") +
  coord_fixed() +
  theme_bw() +
  facet_wrap(~p, labeller = labeller(name = lbl)) +
  guides(color = FALSE) +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank())

In this work we address the piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05]. Tours show interpolated sequences of low-dimensional projections of the data. When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle. However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points. To counteract the effect, we introduce a radial transformation that redistributes data points after projecting them onto a two-dimensional plane, essentially zooming in near the center of the distribution.

XXXX needs to be updated
The paper is structured as follows: the radial transformation is described in Section \ref{sec:method}, the implementation as a new display function for the tourr package is described in Section \ref{sec:implementation}, and we show applications in Section \ref{sec:application}. We conclude in Section \ref{sec:concl}.

XXX probably remove this, for now keeping it in for reference
see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".

@@

XXX show density contours, similar to Fig. 3 - do this as a simulation, move to introduction

@@

```{r radii, fig.cap="Relation between $r$ and $r'$ for different values of $p$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

@@

```{r circles, fig.cap="Rescaled equidistant concentric circles, for $p=2, 3, 10, 100$. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases."}

@@

```{r scaled, eval=F, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$. We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
# define index as function of c
trans_s <- function(s, p){
  p <- s * p
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

@@

XXX this should explain how the display integrates with the tour, in terms of a data pipeline: p-D data -> projection matrix from the tour -> d-D projected data -> radial transformation -> final display

XXX include short tour intro here with relevant references

XXX mention function, but full code should be in documentation

The sage display has been implemented in R [@rref] in the `display_sage` method in the `tourr` package [@tourr]. This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`. Both these parameters are described in Section \ref{sec:params} above.

@@

Below we show example code demonstrates how the new display redistributes the projected points drawn uniformly inside a 5D hypersphere.
```{r codeExampel, echo=TRUE, eval=FALSE}
# use geozoo to uniformly generate points inside a 5D sphere
sphere5 <- geozoo::sphere.solid.random(5)$points
colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5") # naming variables
# standard tour animation with default settings
animate_xy(sphere5)
# new sage display to redistribute the points based on projected volume
animate_sage(sphere5)
# we can use R and s to tune the display
# for example use R to zoom in near the center
# we also adjust half_range to fit the display on the plotting canvas
animate_sage(sphere5, R = 0.1, half_range = 0.1)
```
